Reparametrization of COM-Poisson Regression Models with Applications in the Analysis of Experimental Data

### 2018-01-28 article.Rnw
Add orthogonality study to the article

In this section we performed a simulation study to assess the properties
of the maximum likelihood estimators as well as the flexibility of the
COM-Poisson regression model to deal with non-equidispersed count data.

@@

@@

The results presented in \autoref{tab:coef-cotton} show that the
goodness-of-fit measures (log-likelihood, AIC and BIC) are quite similar
for the COM-Poisson and COM-Poisson$_\mu$ models.  It suggests that the
reparametrization does not change the model fit, as expected.  The
Poisson model is clearly unsuitable, being overly conservative.  The
difference in terms of log-likelihood value from the Poisson to the
COM-Poisson$_\mu$ model was $$\Sexpr{lrt.ncap[2, 4]}$$, which in turn
suggests the better fit of the COM-Poisson$_{\mu}$ model. A chi-square
test also supports this statement.  Finally, the estimated value of the
dispersion parameter $\\hat{\phi} = \Sexpr{co1[1]}$ suggests
underdispersion.

>>>>

In this section we performed a simulation study to assess the properties
of the maximum likelihood estimators and orthogonality of the
reparametrized model as well as the flexibility of the COM-Poisson
regression model to deal with non-equidispersed count data.

@@

To check the orthogonality property we compute the covariance matrix
between maximum likelihood estimators
$\hat{\bm{\theta}} = (\hat{\bm{\beta}}, \phi)$, obtained by observed
information matrix,
Cov$(\hat{\bm{\theta}}) = \mathcal{I}^{-1}(\bm{\theta})$.
\autoref{fig:ortho-plot} shows the covariance between regression and
dispersion parameter estimators for each simulation scenario, on
the correlation scale. The correlations are close to zero in all cases
suggesting the orthogonality property for the reparametrized
model. Interesting results in the first panel shows that
cov($\hat{\beta}_{22}, \hat{\phi}$) is not very close to zero (values
between $-0.4$ and $0.2$) for strong overdispersion ($\phi = -1.6$).

<<ortho-plot, fig.height=4.5, fig.width=9, fig.cap="Empirical correlations between regression and dispersion parameters by different sample sizes and dispersion levels.">>=

#-----------------------------------------------------------------------
# Obtain the covariance between dispersion and regression parameters

aux <- ldply(
    lapply(results, function(x) {
    ind <- names(x); names(ind) <- ind
    out <- lapply(ind, function(y) {
        vcovs <- lapply(x[[y]], "[[", "vcov")
        index <- vapply(lapply(vcovs, is.na), sum, integer(1))
        vcovs <- vcovs[!index]
        corrs <- lapply(vcovs, cov2cor)
        covs <- do.call(rbind, lapply(vcovs, "[", 1, 2:5))
        cors <- do.call(rbind, lapply(corrs, "[", 1, 2:5))
        ldply(list("cov" = covs, "cor" = cors), .id = "scale")
    })
    ldply(out, .id = "phi")
}), .id = "n")
str(aux)

# Organize the results
dacov <- gather(aux, param, value, b0:b22, factor_key = TRUE)
dacov$phi <- ordered(dacov$phi,
                     c("phi=-1.6", "phi=-1", "phi=0", "phi=1.8"))
str(dacov)

# Distributions of covariance/correlations
fl <- parse(text = gsub("=", "==", levels(dacov$phi)))
yl <- parse(text = paste0("hat(beta)[", c(0, 1, 21, 22), "]"))

# Boxplots
bwplot(param ~ value | phi,
       groups = n,
       data = subset(dacov, scale == "cor"),
       box.width = 0.12,
       pch = "|",
       grid = TRUE,
       as.table = TRUE,
       layout = c(4, 1),
       xlab = "Empirical correlations with dispersion parameter estimator",
       fill = colorRampPalette(c("gray80",  "gray20"))(4),
       panel = panel.superpose,
       scales = list(y = list(labels = yl), x = "free"),
       whisker.width = 0.4,
       strip = strip.custom(factor.levels = fl),
       auto.key = list(
           column = 4,
           rectangles = TRUE,
           points = FALSE,
           title = "Sample size",
           cex.title = 1
       ),
       par.settings = list(
           plot.symbol = list(pch = 19, cex = 0.1)
       ),
       panel.groups = function(x, y, group.number, ...) {
           my.panel.bwplot(x, y + (group.number - 2.5) / 5, ...)
           panel.abline(v = 0, lty = 2)
       })

To ilustrate the orthogonality, \autoref{fig:ortho-surf} the deviance
surfaces for four simulated data set of size 1000, $\mu=5$ and different
values of the dispersion parameters. The shapes of the deviance function
show that the proposed parametrization is better for both computation
and asymptotic (normal-based) inference. Furthermore, it is interesting
to note that the deviance function shape on strong overdispersion
($\phi=-1.8$) not as well behaved as the others. This is due to the
difficulty of the distribution in dealing with strong overdispersion in
low counts (see dispersion index plot in the
\autoref{fig:indexes-plot}). This also explains the results of
Cov$(\hat{\beta}_{22}, \phi)$ in the first panel of
\autoref{fig:ortho-plot}, since $\beta_{22}$ is negative and associate
with low counts.

<<ortho-surf, fig.height=5.1, fig.width=9, fig.cap="Deviance surfaces under original and proposed parametrization for simulated data sets of size 1000 an different dispersion parameters. The ellipses are confidence regions (90, 95 and 99\\%), dotted lines are the maximum likelihood estimates, and points are the real parameters used in the simulation.">>=

#-------------------------------------------
# Plots deviance surfaces
output <- readRDS("codes/orthogonality.rds")
devs_orpar <- output$devs_orpar
devs_repar <- output$devs_repar

niveis <- c(0.9, 0.95, 0.99)
cortes <- qchisq(niveis, df = 2)
fl <- parse(text = paste("phi==", sort(phis)))

# Original parametrization
devs_orpar$fphi <- ordered(devs_orpar$realphi)
xy1 <- useOuterStrips(
    levelplot(
        deviance ~ lambda + phi | fphi + "Original parametrization",
        data = devs_orpar,
        scales = list(y = list(rot = 90, relation = "free"),
                      x = "free"),
        cuts = 30,
        xlab = expression(lambda),
        ylab = expression(phi),
        colorkey = FALSE,
        panel = function(x, y, z, at, region, ...,
                         subscripts = subscripts) {
            ##
            flambda <- devs_orpar$fitlambda[subscripts]
            fphi <- devs_orpar$fitphi[subscripts]
            ##
            tlambda <- devs_orpar$reallambda[subscripts]
            tphi <- devs_orpar$realphi[subscripts]
            ##
            panel.levelplot(x, y, z, at = at, region = TRUE,
                            ..., subscripts = subscripts)
            panel.contourplot(x, y, z, ..., at = cortes,
                              contour = TRUE, region = FALSE,
                              subscripts = subscripts)
            panel.abline(v = flambda, h = fphi, lty = 2)
            panel.points(x = tlambda, y = tphi,
                         lty = 2, pch = 19)
        }),
    strip = strip.custom(factor.levels = fl),
    strip.left = strip.custom(par.strip.text=list(cex = 0.9))
)

# Propose parametrization
devs_repar$fphi <- ordered(devs_repar$realphi)
xy2 <- useOuterStrips(
    levelplot(
        deviance ~ mu + phi2 | fphi + "Proposed parametrization",
        data = devs_repar,
        scales = list(y = list(rot = 90, relation = "free"),
                      x = "free"),
        cuts = 30,
        xlab = expression(mu),
        ylab = expression(phi),
        colorkey = FALSE,
        panel = function(x, y, z, at, region, ...,
                         subscripts = subscripts) {
            ##
            fmu <- devs_repar$fitmu[subscripts]
            fphi <- devs_repar$fitphi[subscripts]
            ##
            tmu <- devs_repar$realmu[subscripts]
            tphi <- devs_repar$realphi[subscripts]
            ##
            panel.levelplot(x, y, z, at = at, region = TRUE,
                            ..., subscripts = subscripts)
            panel.contourplot(x, y, z, ..., at = cortes,
                              contour = TRUE, region = FALSE,
                              subscripts = subscripts)
            panel.abline(v = fmu, h = fphi, lty = 2)
            panel.points(x = tmu, y = tphi,
                         lty = 2, pch = 19)
        }),
    strip = strip.custom(factor.levels = fl),
    strip.left = strip.custom(par.strip.text=list(cex = 0.9))
)

# Organize plots
print(xy1, position = c(0.0, 0.48, 1.0, 1.0), more = TRUE)
print(xy2, position = c(0.0, 0.0, 1.0, 0.52), more = FALSE)

@@

The results presented in \autoref{tab:coef-cotton} show that the
goodness-of-fit measures (log-likelihood, AIC and BIC) are quite similar
for the COM-Poisson and COM-Poisson$_\mu$ models.  It suggests that the
reparametrization does not change the model fit, as expected.  The
Poisson model is clearly unsuitable, being overly conservative.  The
difference in terms of log-likelihood value from the Poisson to the
COM-Poisson$_\mu$ model was $\Sexpr{lrt.ncap[2, 4]}$, which in turn
suggests the better fit of the COM-Poisson$_{\mu}$ model. A chi-square
test also supports this statement.  Finally, the estimated value of the
dispersion parameter $\\hat{\phi} = \Sexpr{co1[1]}$ suggests
underdispersion.
